---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

> #### **Loading Libraries**

```{r packages, echo=TRUE, warning=FALSE, message=FALSE}

  library(tidyverse)
  library(ggplot2)
  library(lubridate)

```


> #### **Reading dataset and processing data**

```{r loaddata, echo=TRUE, message=FALSE}
# Check if data unzipped and ready for analysis.

                if (file.exists("activity.csv")) {
                  print("Activity Data Loaded")
                } else {
                  unzip("activity.zip")
                  print("Unzipped & Loaded Activity Data")
                }
# Group data by date, remove NAs, and summarize by the total daily steps. 

                activity_dta <-  read.csv("activity.csv") 

           activity_dta$date <- as.Date(activity_dta$date)
                
# Total Steps per day                
                   actByDate <- activity_dta %>% 
                                  na.omit() %>%
                            group_by(date) %>%
                                summarize(daily_steps = sum(steps))
                                
                   print(head(actByDate))

```


> #### **Histogram of the total number of steps taken each day**

```{r hist, warning=FALSE}
# Histogram of total daily steps 

                ggplot(data = actByDate,
                      aes(x = daily_steps)) +
                      geom_histogram(binwidth = 3000, 
                                     fill = "seagreen",
                                     color = "orange") +
                                     xlab("Daily Steps")+
                                     ylab("Frequency")+
                                     labs(title = "Frequency of total daily steps")+
                                     theme(plot.title = element_text(hjust = 0.5))  

```

> #### **Mean and median number of steps taken each day**

```{r meanactivity, echo=TRUE}

                paste0("mean total number of steps taken per day =  ", round(mean(actByDate$daily_steps),2))

                paste0("average number of steps taken per day =  ", median(actByDate$daily_steps))

```

> #### **Average Daily activity pattern**

```{r avg_pattern, message=FALSE}

# Group data by interval, remove NAs, and summarize by the mean of daily steps.
                
                actByInterval <- activity_dta %>% 
                                    na.omit() %>%
                              group_by(interval) %>%
                                        summarize(daily_avge = mean(steps))
                                       
                                 print(head(actByInterval))
                                        
# Time series of average daily steps      
# Format the interval to 4 char                 
                actByInterval$interval <- sprintf("%04d", actByInterval$interval)
# convert to time
                actByInterval$interval <- as.character(format(strptime(actByInterval$interval,
                                                       format="%H%M"), format = "%H:%M"))%>%
                                          parse_time(actByInterval$interval, format = "%H:%M")
                
# Plot show the time of the day on X and the average daily on Y.
                
                # Define the highest average interval
                
                max_row <- filter(actByInterval, daily_avge == max(daily_avge))
                max_avge_time <- max_row[1] ; max_avge <-  max_row[2]
 
               ggplot(actByInterval,
                      aes(x=`interval`)) +
                      geom_line(aes(y=daily_avge))+
                      xlab("Time of the day(Interval)")+ylab("Daily Average")+
                      labs(title = "Average daily steps by interval")+
                      theme(plot.title = element_text(hjust = 0.5))+
                      geom_point(data=filter(actByInterval, daily_avge == max(daily_avge)),
                                 aes(x = `interval`, y = daily_avge), color = "red", size = 5) 

```

> #### **Interval Contains the maximum average steps.**

```{r maxavge}

                      print(as.data.frame(max_row))

```

> #### **Code to describe and show a strategy for imputing missing data**

```{r NAValues}
                      sum(is.na(activity_dta))

```

> #### **Are there differences in activity patterns between weekdays and weekends?**

```{r weeksplit}
# split weekend and weekdays activities

                  actByDay <- activity_dta 
                  actByDay$Type <- 0
                  
                  for(i in 1:length(actByDay$date)) {
                     actByDay$Type[i] <- ifelse(weekdays(as.Date(actByDay$date[i])) %in% 
                                                        c("Saturday", "Sunday"), "WeekEnd", "WeekDay")
                                  }
                  
                
                actByDay <- actByDay %>% 
                                    na.omit() %>%
                              group_by(interval, Type) %>%
                                        summarize(daily_avge = mean(steps)) 
                
                ggplot(actByDay,
                      aes(x=`interval`)) +
                      geom_line(aes(y=daily_avge)) +
                      facet_grid(Type ~ .)
              
```
