---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

> #### **Loading Libraries**

```{r packages, echo=TRUE, warning=FALSE, message=FALSE}
    
      library(tidyverse)
      library(ggplot2)
      library(lubridate)
    
```


> #### **Load and Process Assignment Data **

```{r loaddata, echo=TRUE, message=FALSE}

# Check if data unzipped and ready for analysis.

if (file.exists("activity.csv")) {
  
                  print("Activity Data Loaded")
                } else {
                  unzip("activity.zip")
                  print("Unzipped & Loaded Activity Data")
                  
                }

devices_data  <-  read.csv("activity.csv") 

devices_data$date <- as.Date(devices_data$date)
                
           
```          
> #### **What is mean total number of steps taken per day?**

> #### **1-Total Daily Steps** 

```{r DS, echo=TRUE, warning=FALSE, message=TRUE}

              # Group data by date, remove NAs, and summarize by the total daily steps.
                
              DS <- devices_data %>% 
                      na.omit() %>%
                      group_by(date) %>%
                      summarize(daily_steps = sum(steps))
                      
                      head(DS)

```

> #### **2-Histogram of the total number of steps taken each day**

```{r hist, warning=FALSE}
# Histogram of total daily steps 

                ggplot(data = DS,
                      aes(x = daily_steps)) +
                      geom_histogram(binwidth = 3000, 
                                     fill = "seagreen",
                                     color = "orange") +
                                     xlab("Daily Steps")+
                                     ylab("Frequency")+
                                     labs(title = "Frequency of total daily steps")+
                                     theme(plot.title = element_text(hjust = 0.5))  

```

> #### **3-Mean and median number of steps taken each day**

```{r meanactivity, echo=TRUE}


                paste0("mean total number of steps taken per day =  ", round(mean(DS$daily_steps),2))
                
 
                paste0("average number of steps taken per day =  ", median(DS$daily_steps))

```

> #### **What is the average daily activity pattern?**

> #### **1-Plot show the time of the day on X and the average daily on Y.**

```{r avg_pattern, message=FALSE}

# Group data by interval, remove NAs, and summarize by the mean of daily steps.
                
                IS <- devices_data %>% 
                                    na.omit() %>%
                              group_by(interval) %>%
                                        summarize(daily_avge = mean(steps))
                                       
                                 head(IS)
                                 
# Format the interval to 4 char to convert to time   
                                 
                IS$interval <- sprintf("%04d", IS$interval)
                
# convert to time 
                IS$interval <- as.character(format(strptime(IS$interval,
                                                       format="%H%M"), format = "%H:%M"))%>%
                                          parse_time(IS$interval, format = "%H:%M")
                
                
# Plot the data to show average daily activity pattern across 5 minutes intervals & highlight the highest average 5 minutes.
                
                # Define the highest average interval
                
                max_row <- filter(IS, daily_avge == max(daily_avge))
                max_avge_time <- max_row[1] ; max_avge <-  max_row[2]
 
               ggplot(IS,
                      aes(x=`interval`)) +
                      geom_line(aes(y=daily_avge))+
                      xlab("Time of the day(Interval)")+ylab("Daily Average")+
                      labs(title = "Average daily steps by interval")+
                      theme(plot.title = element_text(hjust = 0.5))+
                      geom_point(data=filter(IS, daily_avge == max(daily_avge)),
                                 aes(x = `interval`, y = daily_avge), color = "red", size = 5) 

```

> #### **2-Interval Contains the maximum average steps.**

```{r maxavge}

                      print(as.data.frame(max_row))

```

> #### **1 - Calculate and report the total number of missing values in the dataset**

```{r NAs, echo=TRUE}

NAs <- filter(devices_data, is.na(devices_data$steps))
paste0(nrow(NAs), " Missing Value")

```

> #### **Strategy to fill missing Value**

```{r replaceNAs, echo=TRUE}

clean_data  <- devices_data
clean_data[is.na(devices_data)]= mean(devices_data$steps)

DailySteps <- clean_data %>% group_by(date) %>%
                      summarize(daily_steps = sum(steps))

```

> #### ****

```{r hist2, echo=TRUE}

                ggplot(data = DailySteps,
                      aes(x = daily_steps)) +
                      geom_histogram(binwidth = 3000, 
                                     fill = "seagreen",
                                     color = "orange") +
                                     xlab("Daily Steps")+
                                     ylab("Frequency")+
                                     labs(title = "Frequency of total daily steps")+
                                     theme(plot.title = element_text(hjust = 0.5))  
```